### 2022-02-08 symposium.tex
Type and Null Safety for Regexes

The Scala standard library includes a package to manipulate regular expressions.
This package is based on Java's implementation of regular expression, and thus benefits from the high performance of the JVM's regex engine.
The Scala implementation innovates with its presentation: it uses the power and flexibility of Scala's pattern matching to offer a syntax that is both elegant and concise.
The package's documentation starts with the following example:

@@

\noindent
The extractor pattern, |case date(y,m,d) =>|, replaces the need for manually indexing into the regular expression's capturing groups, and shields its users from off-by-one error.

While the syntax of this example would undoubtedly make some non-Scala programmers envious, its type safety leaves much to be desired.
First of all, the \emph{number} of variable bindings in the extractor is entirely opaque to Scala's type system, and left to the discretion of the programmer.
Furthermore, the values coming out of capturing groups can be null (when using optional captures) and thus require an additional layer of validation.
Those shortcomings might appear benign when looking at small examples, but can easily turn into bugs when dealing with a large-scale code base.

In this chapter, we propose a new design for Scala's regular expression library which provides both type-safe and null-safe capturing group extraction.
Our design makes extensive use of match types, Scala~3 new feature for type-level programming, to statically analyze regular expressions during type checking and generate type-specialized extractors.
We build our interface to mimic Scala's original regular expression API so that Scala programmers can use it as a drop-in replacement while still enjoying additional safety.

@@

  \item The term-level capturing group processing, which extracts and sanitizes the output of the regex engine, in accordance with the representation computed by the type-level component (\Cref{sec:term-level}).

  \item The user interface, which ties everything together to compile, execute, and extract the results of a regular expression, while providing type-safe and null-safe capturing group extraction.
\end{enumerate}

@@

The |apply| method is the entry point of our library.
It takes a regular expression and returns an instance of |Regex|.
Instead of directly accepting |String| argument, that method requires a type parameter, |R|, which will be inferred by the compiler, and allows us to get our hands on a type-level instance of the regular expression (the |Singleton| bound is necessary to guide Scala's type inference).
We use that type-level string immediately to instantiate |Compile|, which is our type-level capturing group analysis component.
We store the output of that analysis as |Regex|'s type argument.

The implementation of the |Regex| class is straightforward.
Similar to its standard library counterpart, that class uses an instance of |java.util.regex.Pattern| to match the given input against the regular expression, and to extract the capturing groups whenever the matching succeeds.
The |Regex| class defines an |unapply| method, which the Scala compiler will invoke when using an instance of that class as a pattern matching extractor \citep{emir2007matching}.

@@

\subsection{Capture Group Identification}

@@

Establishing the nullability of capturing groups is more complicated than it sounds.
At first glance, it seems that this is simply a matter of looking for regex quantifiers in suffix position\footnote{
For brevity, our presentation does not account for the regex alternative operator, which also influences the nullability of capturing groups and can appear in both prefix and suffix positions.
In our implementation, we do account for that operator, which we handle analogously.
}.
For instance, in the following naive implementation of |IsNullable|, we look for the first closing parenthesis and inspect the following characters to determine if a capturing group is nullable:

@@

\noindent
Unfortunately, this naive solution is incorrect for regular expressions with nested capturing groups.
For example, in |(A(B)?)|, that solution incorrectly labels the first group as optional.
To overcome this problem, we update our solution to keep track of the number of opening and closing parentheses (|Lvl|), which allows us to differentiate closing parentheses of inner groups from the closing parenthesis of the outer group:

@@

The final iteration of our type-level program handles nested groups by operating in two different modes.
While iterating through the regex's characters, our algorithm can either be outside of an optional group (|Lvl=0|), or inside an option group (|Lvl>0|), in which case it overrides the nullability of newly encountered groups.
Our implementation is as follows (|IsCapturing| omitted for brevity):

\regexLastIteration
% type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Lvl <: Int] =
%   Lo match
%     case Hi => Acc
%     case _  => CharAt[R, Lo] match
%       case ")" => Loop[R, Lo + 1, Hi, Acc, Max[0, Lvl - 1]]
%       case "(" =>
%         Lvl match
%           case 0 => IsNullable[R, Lo + 1, Hi, 0] match
%             case true  => Loop[R, Lo + 1, Hi, Option[String] *: Acc, 1]
%             case false => Loop[R, Lo + 1, Hi, Acc, 0]
%           case _ => Loop[R, Lo + 1, Hi, Option[String] *: Acc, Lvl + 1]
%       case "\\" => Loop[R, Lo + 2, Hi, Acc, Lvl]
%       case _ => Loop[R, Lo + 1, Hi, Acc, Lvl]

>>>>

The Scala standard library contains a package to manipulate regular expressions.
This package is based on Java's implementation, and thus benefits from the high performance of the JVM's regex engine.
The Scala implementation innovates in its presentation: it uses the power and flexibility of Scala's pattern matching to offer a syntax that is both elegant and concise.
The package's documentation starts with the following example:

@@

\noindent
The extractor pattern, |case date(y,m,d) =>|, replaces the need for manually indexing into the regular expression's capturing groups, and shields users from off-by-one error.

While the syntax of this example would undoubtedly make some non-Scala programmers envious, its type safety leaves much to be desired.
First of all, the \emph{number} of variable bindings in the extractor is entirely opaque to Scala's type system, and left to the discretion of the programmer.
Furthermore, the values that come out of capturing groups can be null (when using optional captures) and thus require an additional layer of validation.
Those shortcomings might appear benign on small examples, but can easily turn into bugs when dealing with a large-scale code base.

In this chapter, we propose a new design for Scala's regular expression library which provides a type-safe and null-safe mechanism for capturing group extraction.
Our design makes extensive use of match types, Scala~3 new feature for type-level programming, to statically analyze regular expressions during type checking and generate type-specialized extractors.
We build our interface to mimic Scala's original regular expression API so that Scala programmers can use it as a drop-in replacement and enjoy the additional safety with minimal migration costs.

@@

  \item The term-level capturing group processing, which extracts and sanitizes the output of the regex engine, in accordance with the representation computed at the type level (\Cref{sec:term-level}).

  \item The user interface, which ties everything together to compile, execute, and extract the results of a regular expression, while providing type-safe and null-safe experience.
\end{enumerate}

@@

The |apply| method is the entry point of our library.
It takes a regular expression and returns an instance of |Regex|.
Instead of directly accepting a |String| argument, that method takes a type parameter, |R|, which will be inferred by the compiler, and allows us to get our hands on a type-level instance of the regular expression's string (the |Singleton| bound is necessary to guide Scala's type inference).
We use that type-level string to instantiate |Compile|, our type-level analysis component.

The implementation of the |Regex| class is straightforward.
Similar to its standard library counterpart, that class uses an instance of |java.util.regex.Pattern| to match the given input against its regular expression, and to extract the capturing groups when the matching succeeds.
The |Regex| class defines an |unapply| method, which the Scala compiler invokes when using instances of that class as pattern matching extractors \citep{emir2007matching}.

@@

\subsection{Capturing Group Identification}

@@

Establishing the nullability of capturing groups is more complicated than it sounds.
At first glance, it seems that this is simply a matter of looking for regex quantifiers in suffix position\footnote{
For brevity, our presentation does not account for the regex alternative operator, which also influences the nullability of capturing groups and can appear in both prefix and suffix positions.
Similarly, we omit handling the "at least n times" quantifiers which can lead to nullable capturing groups.
Our implementation accounts for both operators.
}.
For instance, in the following naive implementation of |IsNullable|, we look for the first closing parenthesis and inspect the following characters to determine if a capturing group is nullable:

@@

\noindent
Unfortunately, this naive solution is incorrect for regular expressions with nested capturing groups.
For example, in |(A(B)?)|, that solution incorrectly labels the first group as optional.
To overcome this problem, we update our solution to keep track of the number of opening and closing parentheses (|Lvl|), which allows us to differentiate closing parentheses of inner and outer groups:

@@

We handle nested groups by updated our algorithm to operate in two different modes while iterating through the regex's characters.
The algorithm can either be outside of an optional group (|Opt=0|), in which case it computes nullability of newly encountered via |IsNullable|, or inside an optional group (|Opt>0|), in which case it treats every group as nullable.
This last improvement leads to the final iteration of our type-level program handles nested groups by operating in two different modes.

With this latest improvement, we develop our implementation into its final iteration (|IsCapturing| omitted for brevity):

\regexLastIteration
% type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Opt <: Int] =
%   Lo match
%     case Hi => Acc
%     case _  => CharAt[R, Lo] match
%       case ")" => Loop[R, Lo + 1, Hi, Acc, Max[0, Opt - 1]]
%       case "(" =>
%         Opt match
%           case 0 => IsNullable[R, Lo + 1, Hi, 0] match
%             case true  => Loop[R, Lo + 1, Hi, Option[String] *: Acc, 1]
%             case false => Loop[R, Lo + 1, Hi, Acc, 0]
%           case _ => Loop[R, Lo + 1, Hi, Option[String] *: Acc, Opt + 1]
%       case "\\" => Loop[R, Lo + 2, Hi, Acc, Opt]
%       case _ => Loop[R, Lo + 1, Hi, Acc, Opt]

This concludes the development of our nullability analysis and completes the presentation of the type-level component of our library.
Despite the scarcity of Scala's standard library at the type level, we managed to achieve our ends while keeping our implementation relatively concise and, hopefully, understandable.
In its final iteration, our capturing group analysis is, to the best of our knowledge, on par with Java's implementation of regular expression.
